Supplementary material for  \nBayesian multi-proxy reconstruction of the early Eocene latitudinal temperature gradient

### 2023-02-12 manuscript/manuscript.Rmd
SM fig update

## Early Eocene climate
The geochemical proxy record and ecological data agree that the latitudinal SST gradient of the early Eocene climatic optimum was significantly shallower than the modern [@Huber2011], but beyond that, there is little agreement. @Evans2018 estimate the SST gradient at 32+-10% of the modern, which would correspond to ~`r round( as.numeric(modern_gradient$"50%") - 0.32*as.numeric(modern_gradient$"50%"),0)`+- `r format(round( 0.10*as.numeric(modern_gradient$"50%"),0),nsmall=0)`degC, using the difference between the mean tropical and deep ocean data. The modeled SST of @Tierney2017 using TEX86 is ~ `r round(35.8-23.8,0)` degC, when taking the difference between temperatures at the polar circles and the equator. @Inglis2020 include both terrestrial air temperature and SST estimates in their model of the latitudinal temperature gradient in their Figure S4a; they model a polar circle to equatorial gradient of ~ `r round(30.7-(20.2+14.3)/2,0)`degC. All of those estimates are significantly lower than the SST gradient predicted by an EECO climate model ensemble (@Lunt2012, as shown in @Tierney2017), which predicts a polar circle to equatorial gradient of  `r round(31.5-5.5,0)`degC. Our polar circle to equatorial gradient is the lowest at `r eocene_gradient_polar_circle$"50%"`degC, although the 95% credible interval overlaps with the TEX86-based estimates of @Tierney2017 and the composite estimate of @Inglis2020.

@@

In latitudes beyond the polar circle, the discrepancy between our model predictions and those of earlier, proxy-based models increases, as those predict almost linearly decreasing SSTs towards the poles, whereas our median model prediction suggests only a slight decrease beyond the polar circle. However, the scarcity of temperature records in this range leads to widening credible intervals, including the possibility of stronger temperature decreases. Polar temperature estimates from our model are thus conservative in that they admit large uncertainty where data is absent, which is desirable. However, the presence of high proxy-derived temperature estimates at ~ 60 deg latitudes forces the modeled median temperature curve to be too high at ~ 24 degC, relative to the temperatures indicated by the high-latitude mangrove communities (15.6 - 22.5 degC). The very high variability of the proxy record in mid-latitudes results in large uncertainties on the shape of temperature gradient and on the GMSsT. Biases and errors in the proxy reconstructions likely contribute to the observed variability, as geochemical proxies reflect many other factors besides seawater temperature [@Hollis2019]. Despite excluding d18O measurements from recrystallised fossils, systematic offsets remain between mostly warm temperatures derived from TEX86, and cooler temperatures derived from d18O, delta47, and the ecological proxies. Seasonality [@Keating-Bitoni] and temporal changes within the EECO [@Westerhold2018] may also contribute to the large variability of the EECO proxy data.

>>>>

## Early Eocene climate
The geochemical proxy record and ecological data agree that the latitudinal SST gradient of the early Eocene climatic optimum was significantly shallower than the modern [@Huber2011], but beyond that, there is little agreement. @Evans2018 estimate the SST gradient at 32+-10% of the modern, which would correspond to ~`r round( as.numeric(modern_gradient$"50%") - 0.32*as.numeric(modern_gradient$"50%"),0)`+- `r format(round( 0.10*as.numeric(modern_gradient$"50%"),0),nsmall=0)`degC, using the difference between the mean tropical and deep ocean data. The modeled SST of @Tierney2017 using TEX86 is ~ `r round(35.8-23.8,0)` degC, when taking the difference between temperatures at the polar circles and the equator. @Inglis2020 include both terrestrial air temperature and SST estimates in their model of the latitudinal temperature gradient in their Figure S4a; they model a polar circle to equatorial gradient of ~ `r round(30.7-(20.2+14.3)/2,0)`degC. All of those estimates are significantly lower than the SST gradient predicted by an EECO climate model ensemble (@Lunt2012, as shown in @Tierney2017), which predicts a polar circle to equatorial gradient of  `r round(31.5-5.5,0)`degC. @Keating-Bitoni show a proxy-based SST gradient of ~ `r round(30-(19+15.6)/2,0)`degC, and @Bij2009 model a gradient of ~`r round(36-27,0)`degC in the northern and only ~`r 36-31`degC in the southern hemisphere.
Our polar circle to equatorial gradient is lower than most previous estimates at `r eocene_gradient_polar_circle$"50%"`degC, although the 95% credible interval overlap with several of the listed gradient estimates.

@@

In latitudes beyond the polar circle, the discrepancy between our model predictions and those of earlier, proxy-based models increases, as those predict almost linearly decreasing SSTs towards the poles, whereas our median model prediction suggests only a slight decrease beyond the polar circle. However, the scarcity of temperature records in this range leads to widening credible intervals, including the possibility of stronger temperature decreases. Polar temperature estimates from our model are thus conservative in that they admit large uncertainty where data is absent, which is desirable. However, the presence of high proxy-derived temperature estimates at ~ 60 deg latitudes forces the modeled median temperature curve to be too high at ~ 24 degC, relative to the temperatures indicated by the high-latitude mangrove communities (15.6 - 22.5 degC). In contrast, the extrapolated polar temperatures of most previous proxy-based models are likely too low, given the abundance of ecological data indicating temperate or subtropical high-latitude climates during the EECO [@ @]

The very high variability of the proxy record in mid-latitudes results in large uncertainties on the shape of temperature gradient and on the GMSsT. Biases and errors in the proxy reconstructions likely contribute to the observed variability, as geochemical proxies reflect many other factors besides seawater temperature [@Hollis2019]. Despite excluding d18O measurements from recrystallised fossils, systematic offsets remain between mostly warm temperatures derived from TEX86, and cooler temperatures derived from d18O, delta47, and the ecological proxies. Seasonality [@Keating-Bitoni] and temporal changes within the EECO [@Westerhold2018] may also contribute to the large variability of the EECO proxy data.

### 2023-02-12 manuscript/supplementary_material.Rmd
SM fig update

```{r out.width='100%', echo = FALSE, fig.cap = "Fig S4. Simulated latitudinal temperature gradients sampled at the absolute latitudes of the EECO data. The dark grey points show the simulated temperature data, the black line shows the shape of the true gradient, and the medians (lines) and 95% credibal intervals (shadings) of the inferred gradients using the model are shown in red. \\\n a) linear gradient; $SST = 30-|latitude|/3$ \\\n b) flat gradient; $SST = 25$ \\\n c) quadratic gradient; $SST = 30-|latitude|^2/270$ \\\n d) logit-shaped gradient; $SST = 14-3*log(|latitude|/(90-|latitude|))$ \\\n  Except for d), which is very unlike the modern gradient or reconstructed gradients, all latitudinal temperature gradients are well captured by the model in the latitudinal range for which data is provided."}
setwd('..') # move up to parent folder of /manuscript
knitr::include_graphics("D://Github/PalaeoClimateGradient/figures/SM/FigS4_different_gradients.png", error = FALSE)
```

>>>>

```{r out.width='100%', echo = FALSE, fig.cap = "Fig S4. Simulated latitudinal temperature gradients sampled at the absolute latitudes of the EECO data. The dark grey points show the simulated temperature data, the black line shows the shape of the true gradient, and the medians (lines) and 95% credibal intervals (shadings) of the inferred gradients using the model are shown in red. \\\n a) linear gradient; $SST = 30-|latitude|/3$ \\\n b) flat gradient; $SST = 25$ \\\n c) quadratic gradient; $SST = 30-|latitude|^2/270$ \\\n d) as c), but the priors on A, M, and B were relaxed to allow for a better fit to the quadratic gradient in high latitudes \\\n  All latitudinal temperature gradients are well captured by the model in the latitudinal range for which data is provided. If the priors, particularly the prior on the lower limit of the gradient (A), are relaxed (d), the model can approximate a quadratic or linear function even in high latitudes and at cold temperatures."}
setwd('..') # move up to parent folder of /manuscript
knitr::include_graphics("D://Github/PalaeoClimateGradient/figures/SM/FigS4_different_gradients.png", error = FALSE)
```
